---
title: "MEFISTO application to spatial transcriptomics data"
author:
- name: "Britta Velten"
  affiliation: "German Cancer Research Center, Heidelberg, Germany"
  email: "b.velten@dkfz-heidelberg.de"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{MEFISTO on spatial transcriptomics data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, warning=FALSE, message=FALSE}
library(MOFA2)
library(tidyverse)
library(cowplot)
library(magrittr)
```

# Load ST data set
We use the example data provided in the [Seurat vignette](https://satijalab.org/seurat/v3.2/spatial_vignette.html), which contains spatial transcriptomics data on a slide of a mouse brain tissue generated using 10x visium.

```{r}
library(Seurat)
SeuratData::InstallData("stxBrain")
brain <- SeuratData::LoadData("stxBrain", type = "anterior1")
brain <- Seurat::NormalizeData(brain, assay = "Spatial", verbose = FALSE)
brain <- Seurat::FindVariableFeatures(brain, assay = "Spatial")
```

# Preprocess data for MEFISTO
Next, we prepare the data for input to MEFISTO. For this we need the expression data  (using the 2000 top variable features) as well as the spatial coordinates of each sample (spot).
```{r}
expression_data <- as.matrix(brain@assays$Spatial@data[VariableFeatures(brain),])

dim(expression_data)
```
```{r}
locs <- GetTissueCoordinates(brain)
locs <- tibble::rownames_to_column(locs, "sample") %>%
  gather(key = "covariate", value = "value", starts_with("image"))

head(locs %>% arrange(sample))
```

# Create a MOFA object specifying spatial coordinates as covariates
To create a space-aware MOFA model (= MEFISTO) we provide MOFA the expression data as well as the spatial coordinates in `set_covariates`.
```{r}
# create MOFA objects and specify the spatial coordinates as covariates
sm <- create_mofa(data = list(RNA = expression_data))
sm <- set_covariates(sm, locs)
```

# Prepare training
Before training, we can specify various options for the model, the training and the data pre-processing. If no options are specified, the model will use the default options. See also `get_default_data_options`, `get_default_model_options`, `get_default_training_options` and `get_default_smooth_options` to have a look at the various options and their defaults. Here, we in particular, specify that we want to use spare Gaussian process with a smaller number of inducing points which can speed up training.
```{r}
model_opts <- get_default_model_options(sm)
model_opts$num_factors <- 4

n_inducing <- 1000 # We use 1000 inducing points to learn spatial covariance patterns
smooth_opts <- get_default_smooth_options(sm)
smooth_opts$sparseGP <- TRUE
smooth_opts$frac_inducing <- n_inducing / get_dimensions(sm)$N

train_opts <- get_default_training_options(sm)
train_opts$seed <- 2021

sm <- prepare_mofa(sm, training_options = train_opts,
                   model_options = model_opts,
                   smooth_options = smooth_opts)
```


```{r}
outfile = "data/ST_model.hdf5"
```

Next, we train the model. As running the model can take some time, we here load a pre-trained model.
```{r, eval = FALSE}
# sm <- run_mofa(sm, outfile = outfile)
```

# Load a trained model
```{r, warning=FALSE}
sm <- load_model(outfile)
```

```{r, echo = FALSE}
# as the model was trained without names we add feature names back on trained model
# this is not required if the model was trained with names or from R with the run_mofa function. 
features_names(sm) <- list(rownames(expression_data))
```


# Downstream analyses
## Variance decomposition
First, we can take a look whether our factor are uncorrelated.
```{r, fig.height=3, fig.width=3}
plot_factor_cor(sm)
```

## Spatial factors
Next, we will have a look at the spatial patterns that are captured by each factor. 
We here rotate the tissue to obtain the same as orientation as in `SpatialFeaturePlot` below.

```{r, fig.width=18, fig.height=4}
plot_factors_vs_cov(sm, rotate_y = TRUE)
```

## Smoothness of factors
All of this factors seem to capture spatial patterns of variation that seems to vary smoothly along space to some extent. We can take a look at the smoothness score inferred by the model.
```{r, fig.height=0.5, fig.width=7}
plot_smoothness(sm)
```

## Weights
To inspect which genes underly these spatial patterns, we can take a look at the top weights per factor. For example, Factor 4 highlights Ttr, a marker of the choroid plexus.
```{r, fig.width=3, fig.height=4}
plot_top_weights(sm, factors = 4)
```

To explore the expression patterns of these genes that are highlighted on the weights by MEFISTO, we can use Seurat's `SpatialFeaturePlot`.
```{r}
W <- get_weights(sm)[[1]]
top_weights <- rownames(W)[apply(abs(W), 2, which.max )]
list_seurat <- SpatialFeaturePlot(brain, features = top_weights,
                                combine = F)
gg_seurat <- plot_grid(plotlist = list_seurat, nrow = 1)
gg_seurat
```


# SessionInfo
```{r}
sessionInfo()
```

